---
title: "2D Finite Difference Model"
author: "Claire Harrigan"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: html_document
---

```{r setup, include = FALSE}
# This RMarkdown file creates a PDF of a finite difference model.

# Clear all from workspace/environment.
  rm(list = ls())

# Set universal settings for all code chunks.
  knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE) 
  
# Install and load packages.
  # install.packages("ggplot2")
  library(ggplot2)
```

```{r ModelDimensions}
# Define size of model domain.
  x.size <- 100000  # [m] 
  z.size <- 100000  # [m] 

# Set resolution.  
  Nx <- 40
  Nz <- 40
  
# Define stepsize.
  # dx <- 1000  # [m] 
  # dz <- 1000  # [m]
  dx <- x.size / (Nx - 1)  # [m] 
  dz <- z.size / (Nz - 1)  # [m]  

# Calculate number of nodes.
  x.num <- (x.size / dx) + 1
  z.num <- (z.size / dz) + 1

# Create vectors of node locations.
  x <- seq(from = 0, to = x.size, by = dx)  # [m]
  z <- seq(from = 0, to = z.size, by = dz)  # [m]
```

```{r MaterialProperties}
# Define initial material properties.
  A <- 0  # heat production, [W/(m^3)] (1e-6)
  k <- 3  # thermal conductivity, [W/(m*K)]
  U <- 0  # uplift, # [m/s]
  rho <- 3300  # density, [kg/(m^3)]
  Cp <- 1000  # heat capacity, [J/(kg*K)]

# Calculate minimum thermal diffusivity value to calcualte a stable time step.
  Kappa.min <- k / (rho * Cp)  # [(m^2)/s]

# Define time step.
  # dt <- (dx ^ 2) / 6 * Kappa.min  # [s]
  dt <- (dx ^ 2) / (3 * Kappa.min)  # [s]
```

```{r InitializeArrays}
# Initialize a matrix for initial temperature conditions.
  T.0 <- matrix(data = NA,
               nrow = z.num,
               ncol = x.num)

# Initialize a matrix to hold temperature values conditions.
  T.dt <- matrix(data = NA,
                 nrow = z.num,
                 ncol = x.num)

# Initialize a matrix of thermal diffusivity values.
  Kappa <- matrix(data = Kappa.min,
                  nrow = z.num,
                  ncol = x.num)

# Intialize a matrix of heat production values.
  A.0 <- matrix(data = A,
                nrow = z.num,
                ncol = x.num)
```

```{r InitializeTDistribution_plume}
# Set up an initial condition of a circular plume at an elevated temperature.

# Set plume radius.
  r.plume <- 1000  # [m]

# Initialize rnode.
  r.node <- matrix(data = NA,
                   nrow = z.num,
                   ncol = x.num)

# Loop through the nodes and assign a temperature to the initial conditions
# based on whether a node is within the plume.  
  for (i in 1:x.num) {
    for (j in 1:z.num) {
      r.node <- ((x[[i]] - x.size / 2) ^ 2 + (z[[j]] - z.size / 2) ^ 2) ^ 0.5
      if (r.node > r.plume) {
        T.0[i, j] = 1500 + 273.15
      } else {
        T.0[i, j] = 1700 + 273.15
      }
    }
  }
  
# Clean up.
  rm(i, j)
```

### Initial Condtions  
  
**Model Parameters:**  
dt = `r dt` s  
dx = `r dx` m  
dz = `r dz` m  

**Material Properties:**  
Heat production, A = `r A` W/(m^3)   
Thermal conductivity, k = `r k` W/(m K)  
Density, rho = `r rho` kg/(m^3)  
Heat capcity, Cp = `r Cp` J/(kg K)  
Uplift, U = `r U` m/s  

**Initial perterbation:**  
circular plume with `r r.plume` m radius at an elevated temperature  

```{r PlotInitialConditions}
# Plot initial conditions.

# Make a data frame of initial condtions.
  for.plotting <- as.data.frame(T.0)

# Rename columns and rows of data frame (for plotting).
  rownames(for.plotting) <- x
  colnames(for.plotting) <- z

# Load package reshape2 for melting data frames.
  library(reshape2)

# Melt data frame for plotting.
  melted.for.plotting <- melt(for.plotting)

# Load packages plotly for plotting.
  library(plotly)

# Plot initial condtions as a heatmap.
  plot_ly(
    x = x,
    y = z, 
    z = T.0,
    type = "heatmap",
    colorscale = "Viridis"
  )
```

```{r Model_SetIterations}
# Set a number of time steps.
  n <- 1000

# Initialize a list as a container for model returns at different times.
  T.n  <- list()
  
# Place T.0 in the first position of the T.n list.
  T.n[[1]] <- T.0

# Solve the finite difference equation for n time steps.
  
  for (t in 1:n) {
    
    # Initialize T.n[[t + 1]]
    T.n[[t + 1]] <- matrix(data = NA,
                           nrow = z.num,
                           ncol = x.num)
    
    # Set constant T border.
    T.n[[t + 1]][1, ] <- T.n[[t]][1, ]          # top
    T.n[[t + 1]][z.num, ] <- T.n[[t]][z.num, ]  # bottom
    T.n[[t + 1]][, 1] <- T.n[[t]][, 1]          # left
    T.n[[t + 1]][, x.num] <- T.n[[t]][, x.num]  # right
    
    # Calculate T for interior of T.n[[t + 1]] matrix.
    for (i in 2:(x.num - 1)) {
      for (j in 2:(z.num - 1)) {
        
        T.n[[t + 1]][i, j] <- T.n[[t]][i, j] +
          
          # conduction term
          (Kappa.min * dt * (((T.n[[t]][(i + 1), j] - (2 * T.n[[t]][i, j]) + T.n[[t]][(i - 1), j]) / (dx ^ 2)) + ((T.n[[t]][i, (j + 1)] + (2 * T.n[[t]][i, j]) + T.n[[t]][i, (j - 1)]) / (dz ^ 2) ))) +
          
          # heat production term
          ((A.0[i, j] * dt) / (rho * Cp)) +  
          
          # advection/uplift term
          (U * dt * ((T.n[[t]][i, (j + 1)] - T.n[[t]][i, j]) / dz))  
      }
    }
  }

# Check to see the temperature change.
  # T.n[[length(T.n)]] - T.n[[1]]
  # max(T.n[[length(T.n)]] - T.n[[1]])
```

### Results after `r eval(n)` iterations

Maximum temperature change between model run n = `r n` (time = `r n * dt` s) and initial condition (time = 0 s):  
`r  max(T.n[[length(T.n)]] - T.n[[1]])` K  
```{r PlotModel_1}
# Plot model results for a specified time step.
  plot_ly(
    x = x,
    y = z, 
    z = T.n[[length(T.n)]],
    type = "heatmap",
    colorscale = "Viridis"
  )
```


IN PROGRESS

```{r Model_SetIterationsOverwrite, eval=FALSE, message=FALSE, warning=FALSE, include=FALSE}
### Model run without saving every time step

# Set a number of time steps.
  n <- 1000

# Initialize a list as a container for model returns at different times.
  T.n.ow  <- list()
  
# Place T.0 in the first position of the T.n.ow list.
  T.n.ow[[1]] <- T.0

# Solve the finite difference equation for n time steps.
  
  for (t in 1:n) {
    
    # Initialize T.n.ow[[t + 1]]
    T.n.ow[[t + 1]] <- matrix(data = NA,
                           nrow = z.num,
                           ncol = x.num)
    
    # Set constant T border.
    T.n.ow[[t + 1]][1, ] <- T.n.ow[[t]][1, ]          # top
    T.n.ow[[t + 1]][z.num, ] <- T.n.ow[[t]][z.num, ]  # bottom
    T.n.ow[[t + 1]][, 1] <- T.n.ow[[t]][, 1]          # left
    T.n.ow[[t + 1]][, x.num] <- T.n.ow[[t]][, x.num]  # right
    
    # Calculate T for interior of T.n.ow[[t + 1]] matrix.
    for (i in 2:(x.num - 1)) {
      for (j in 2:(z.num - 1)) {
        
        T.n.ow[[t + 1]][i, j] <- T.n.ow[[t]][i, j] +
          
          # conduction term
          (Kappa.min * dt * (((T.n.ow[[t]][(i + 1), j] - (2 * T.n.ow[[t]][i, j]) + T.n.ow[[t]][(i - 1), j]) / (dx ^ 2)) + ((T.n.ow[[t]][i, (j + 1)] + (2 * T.n.ow[[t]][i, j]) + T.n.ow[[t]][i, (j - 1)]) / (dz ^ 2) ))) +
          
          # heat production term
          ((A.0[i, j] * dt) / (rho * Cp)) +  
          
          # advection/uplift term
          (U * dt * ((T.n.ow[[t]][i, (j + 1)] - T.n.ow[[t]][i, j]) / dz))  
      }
    }
  }
```

```{r PlotModel_2}
# Plot model results for a specified time step.
  plot_ly(
    x = x,
    y = z, 
    z = T.n.ow[[length(T.n.ow)]],
    type = "heatmap",
    colorscale = "Viridis"
  )
```











```{r Model_SteadyStateStop, eval=FALSE, message=FALSE, warning=FALSE, include=FALSE, paged.print=TRUE}
# Set a maximum number of time steps.
  n <- 10

# Set a temperature threshold, a value of change in temperature between the same
# nodes on matrices T.n[n] and T.n[n+1] that suggests that the model has reached
# steady state.
  T.threshold <- 2

# Initialize a list as a container for model returns at different times.
  T.n  <- list()

# Place T.0 in the first position of the T.n list.
  T.n[[1]] <- T.0
  
# Make a matrix that is more than T.treshold greater than T.0 so the first 
# iteration of the while loop below can be applied and continue running. This 
# matrix will be overwritten in the for loop below.
  T.n[[2]] <- T.0 + (2 * T.threshold)
  
# Solve the finite difference equation for T.dt for a maximum of n time steps 
# until the T difference between T.n[n] and T.n[n+1] is less than a set 
# temperature threshold. Use constant T edges.
 
  
  # while (T.n[[t + 1]] - T.n[[t]] < T.threshold) {
  #   T.n[[t + 1]] <- T.n[[t]] + 3
  # }

  
  # if (T.n[[t + 1]] - T.n[[t]] < T.threshold) {
  #   print("yay")
  # } else {
  #   print("nay")
  # }
   
  # while (T.n[[t + 1]] - T.n[[t]] < T.threshold) {
  #   for (t in 1:n) {
  #     T.n[[t + 1]] <- T.n[[t]] + 3
  #   }
  # }
  

  
  # Tn1 = Tn +
  #   (Kappa * dt * (((Tn - (2 * Tn) + Tn ) / (dx ^ 2)) + ((Tn + (2 * Tn) + Tn) / (dz ^ 2) ))) + 
  #   ((A.0 * dt) / (rho * Cp)) +
  #   (U * dt * ((Tn - Tn) / dz))

  
  # WORKING CODE  
  
  for (t in 1:n) {
    # If the maximum value of the difference between the last and second to last
    # model results (arguments in list) is less than a threshold value, compute
    # the next model solution.
    # if (max(T.n[[length(T.n)]] - T.n[[(length(T.n) - 1)]]) > T.threshold) {
    
    # Initialize T.n[[t + 1]]
    T.n[[t + 1]] <- matrix(data = NA,
                           nrow = z.num,
                           ncol = x.num)
    
    # Set constant T border.
    T.n[[t + 1]][1, ] <- T.n[[t]][1, ]          # top
    T.n[[t + 1]][z.num, ] <- T.n[[t]][z.num, ]  # bottom
    T.n[[t + 1]][, 1] <- T.n[[t]][, 1]          # left
    T.n[[t + 1]][, x.num] <- T.n[[t]][, x.num]  # right
    
    # Calculate T for interior of T.n[[t + 1]] matrix.
    for (i in 2:(x.num - 1)) {
      for (j in 2:(z.num - 1)) {
        T.n[[t + 1]][i, j] <- T.n[[t]][i, j] + t # Replace with real equation
        #   Tn1 = Tn +
        # (Kappa * dt * (((Tn - (2 * Tn) + Tn ) / (dx ^ 2)) + ((Tn + (2 * Tn) + Tn) / (dz ^ 2) ))) + 
        # ((A.0 * dt) / (rho * Cp)) +
        # (U * dt * ((Tn - Tn) / dz))
      }
    }
    # } 
  }

  

```

```{r Adiabat, eval=FALSE, message=FALSE, warning=FALSE, include=FALSE}
# adiabat (0.5K/km) from 1500 at top to something
```

```{r Notes, eval=FALSE, message=FALSE, warning=FALSE, include=FALSE}

# Indexing
        # T.n[[t]][i, j]
        # T.n[[t]][(i + 1), j]
        # T.n[[t]][(i - 1), j]
        # T.n[[t]][i, (j + 1)]
        # T.n[[t]][i, (j - 1)]
  

# visualize, solve, reset containers 
# loop with replacement

# visualize T0
# for
# solve Tn+1
# visualize/sample Tn+1 -> output "pcolor" (color-coded T grid)
# set T0 = Tn+1
# next

# to find steady state, compare Tn+1 and Tn... if diff < threshold then exit

# initialize n matrices
# for 
# solve Tn
# append to list
# next

# while loop to set up threshold ahead of time

# use wraparound or constant T edges


# ## SOLVE FINITE DIFERENCE MODEL ------------------------------------------------
# 
#   # Populate the last column of the matix.
#   TC[, dim(TC)[2]] <- TC[1, dim(TC)[2]]
# 
#   # Loop though the cells in the matrix and calculate model values.
#   # Initialize vector used in for loop.
#   TCvector <- vector()
#   # Run for loop.
#   for (j in 2:dim(TC)[1]) {  
#     for (i in 2:(dim(TC)[2] - 1)) {
#       TCvector[[1]] <- Ts
#       TCvector[[i]] <- TC[j - 1, i] + 
#         (K * delta.t.s) * (((TC[j - 1, i + 1] - 
#                                (2 * TC[j - 1, i]) + 
#                                TC[j - 1, i - 1]) / (delta.x ^ 2)) 
#                            + (A / k) 
#                            + ((U * (TC[j - 1, i + 1] - 
#                                       TC[j - 1, i])) / (K * delta.x)))
#     }
#     TC[j, c(2:dim(TC)[2] - 1)] <- TCvector
#   }




```

