---
title: "2D Finite Difference Model"
author: "Claire Harrigan"
date: "`r format(Sys.time(), '%d %B %Y')`"
output: html_document
---

```{r setup, include = FALSE}
# This RMarkdown file creates an html file of a finite difference model.

# Clear all from workspace/environment.
  rm(list = ls())

# Set universal settings for all code chunks.
  knitr::opts_chunk$set(echo = FALSE, message = FALSE, warning = FALSE) 
```

```{r ModelDimensions}
# Define size of model domain.
  x.size <- 100000  # [m] 
  z.size <- 100000  # [m] 

# Set resolution (number of nodes).  
  x.num <- 40 
  z.num <- 40 
  
# Define stepsize.
  dx <- x.size / (x.num - 1)  # [m] 
  dz <- z.size / (z.num - 1)  # [m]  

# Create vectors of node locations.
  x <- seq(from = 0, to = x.size, by = dx)  # [m]
  z <- seq(from = 0, to = z.size, by = dz)  # [m]
```

```{r Constants}
# Define material properties, initialize arrays.
  k <- matrix(data = 3.0,      # thermal conductivity, [W/(m*K)]
              nrow = z.num,
              ncol = x.num)  
  A <- matrix(data = 2.2e-8,   # heat production, [W/(m^3)] (1e-6)
              nrow = z.num,
              ncol = x.num)  
  rho <- matrix(data = 3300,   # density, [kg/(m^3)] (2800)
                nrow = z.num,
                ncol = x.num)
  Cp <- matrix(data = 1293,    # heat capacity, [J/(kg*K)] (1000)
               nrow = z.num,
               ncol = x.num)
  Kappa <- k / (rho * Cp)      # [(m^2)/s]
  Kappa.min <- min(Kappa)      # [(m^2)/s] # Kappa.min <- 1e-6   # (Buchanan)
  U <- matrix(data = 0,        # uplift, [m/s]
              nrow = z.num,
              ncol = x.num)
                   
# Define time step.
  dt.max <- (dx ^ 2) / (6 * Kappa.min)  # [s] 
  dt.yr <- dt.max * 3.17098e-8 # [yr]
  dt <- dt.max
```

```{r InitializeTDistribution_plume}
# Set up an initial condition of a circular plume at an elevated temperature.
  
# Initialize a matrix for initial temperature conditions.
  T.0 <- matrix(data = 0,
                nrow = z.num,
                ncol = x.num)

# Set plume radius.
  r.plume <- 20000  # [m]

# Initialize r.node matrix.
  r.node <- matrix(data = NA,
                   nrow = z.num,
                   ncol = x.num)

# Loop through the nodes and assign a temperature to the initial conditions
# based on whether a node is within the plume.  
  for (i in 1:x.num) {
    for (j in 1:z.num) {
      r.node <- ((x[[i]] - x.size / 2) ^ 2 + (z[[j]] - z.size / 2) ^ 2) ^ 0.5
      if (r.node > r.plume) {
        # Set conditions in the mantle.
        T.0[i, j]   <- 1500 + 273.15
        # T.0[i, j]   <- 1500 + (0.5 * -z[[j]] / 1000) + 273.15
        # k[i, j]     <- 3.0    # thermal conductivity, [W/(m*K)]
        # A[i, j]     <- 1e-6   # heat production, [W/(m^3)]
        # rho[i, j]   <- 3300   # density, [kg/(m^3)]
        # Cp[i, j]    <- 1293   # heat capacity, [J/(kg*K)]
        # Kappa[i, j] <- k[i, j] / (rho[i, j] * Cp[i, j])  # [(m^2)/s]
      } else {
        # Set conditions in the plume.
        T.0[i, j]   <- 1700 + 273.15
        # k[i, j]     <- 3.0    # thermal conductivity, [W/(m*K)]
        # A[i, j]     <- 1e-6   # heat production, [W/(m^3)]
        # rho[i, j]   <- 3300   # density, [kg/(m^3)]
        # Cp[i, j]    <- 1293   # heat capacity, [J/(kg*K)]
        # Kappa[i, j] <- k[i, j] / (rho[i, j] * Cp[i, j])  # [(m^2)/s]
      }
    }
  }
  
# Clean up.
  rm(i, j)
```

### Initial Condtions  
  
**Model Parameters:**  
dt = `r round((dt * 3.17098e-8 / 1000000), digits = 3)` Myr)
dx = `r round(dx, digits = 1)` m  
dz = `r round(dz, digits = 1)` m  

**Material Properties:**  
Heat production, A = `r mean(A)` W/(m^3)   
Thermal conductivity, k = `r mean(k)` W/(m K)  
Density, rho = `r mean(rho)` kg/(m^3)  
Heat capcity, Cp = `r mean(Cp)` J/(kg K)  
Uplift, U = `r mean(U)` m/s  

**Initial perterbation:**  
circular intrusion with `r (r.plume / 1000)` km radius at an elevated temperature (1700°C)  

```{r PlotInitialConditions_heatmap}
# Plot initial conditions.

# Load packages plotly and Rcolorbrewer for plotting.
  library(plotly)
  library(RColorBrewer)

# Plot initial condtions as a heatmap.
  plot_ly(x = (x / 1000),
          y = (z / 1000), 
          z = T.0 - 273.15,
          type = "heatmap",
          colors = "Greys") %>%
          # colors = rev(brewer.pal(9,"RdYlBu"))) %>%
    colorbar(title = "Temperature (°C)",
             limits = c(1450, 1750))  %>%
  layout(xaxis = list(title = "Distance (km)"),
         yaxis = list(title = "Depth (km)",
                      autorange = "reversed"))
```

```{r Model_SetIterations}
# Set a number of time steps.
  n <- 100  # 100 iterations is about 5 myr

# Initialize a list as a container for model returns at different times.
  T.n  <- list()
  
# Place T.0 in the first position of the T.n list.
  T.n[[1]] <- T.0

# Solve the finite difference equation for n time steps.
  
  for (t in 1:n) {
    
    # Initialize T.n[[t + 1]]
    T.n[[t + 1]] <- matrix(data = NA,
                           nrow = z.num,
                           ncol = x.num)
    
    # Calculate T for interior of T.n[[t + 1]] matrix.
    for (i in 2:(x.num - 1)) {
      for (j in 2:(z.num - 1)) {
            
        # Set constant T border.
        T.n[[t + 1]][1, ] <- T.n[[t]][1, ]          # top
        T.n[[t + 1]][z.num, ] <- T.n[[t]][z.num, ]  # bottom
        T.n[[t + 1]][, 1] <- T.n[[t]][, 1]          # left
        T.n[[t + 1]][, x.num] <- T.n[[t]][, x.num]  # right
        # T.n[[t + 1]][, 1] <- 1500 + (0.5 * -z[[i]] / 1000) + 273.15      # left
        # T.n[[t + 1]][, x.num] <- 1500 + (0.5 * -z[[i]] / 1000) + 273.15  # right
        
        # Calculate T in the interior of the model
        T.n[[t + 1]][i, j] <- T.n[[t]][i, j] +
          
          # conduction term
          ((Kappa[i, j] * dt) * (((T.n[[t]][(i + 1), j] - (2 * T.n[[t]][i, j]) + T.n[[t]][(i - 1), j]) / (dx ^ 2)) + ((T.n[[t]][i, (j + 1)] - (2 * T.n[[t]][i, j]) + T.n[[t]][i, (j - 1)]) / (dz ^ 2) ))) +
          
          # heat production term
          ((A[i, j] * dt) / (rho[i, j] * Cp[i, j])) +

          # advection/uplift term
          (U[i, j] * dt * ((T.n[[t]][i, (j + 1)] - T.n[[t]][i, j]) / dz))
      }
    }
  }
```

### Results after `r eval(n)` iterations

Maximum temperature change between model run n = `r n` (time = `r round((n * dt * 3.17098e-8 / 1000000), digits = 2)` Myr) and initial condition (time = 0 Myr):  
`r  round(((max(T.n[[length(T.n)]] - T.n[[1]])) - 273.15), digits = 2)` °C   

```{r PlotModel_heatmap}
# Plot model results after n iterations.
  plot_ly(x = (x / 1000),
          y = (z / 1000), 
          z = T.n[[length(T.n)]] - 273.15,
          type = "heatmap",
          colors = "Greys") %>%
          # colors = rev(brewer.pal(9,"RdYlBu"))) %>%
    colorbar(title = "Temperature (°C)",
             limits = c(1450, 1750))  %>%
  layout(xaxis = list(title = "Distance (km)"),
         yaxis = list(title = "Depth (km)",
                      autorange = "reversed"))
```


```{r Model_SetIterations_2}
# Set a number of time steps.
  m <- 100  # 100 iterations is about 5 myr

# Solve the finite difference equation for n time steps.
  
  for (t in n:(n + m)) {
    
    # Initialize T.n[[t + 1]]
    T.n[[t + 1]] <- matrix(data = NA,
                           nrow = z.num,
                           ncol = x.num)
    
    # Calculate T for interior of T.n[[t + 1]] matrix.
    for (i in 2:(x.num - 1)) {
      for (j in 2:(z.num - 1)) {
            
        # Set constant T border.
        T.n[[t + 1]][1, ] <- T.n[[t]][1, ]          # top
        T.n[[t + 1]][z.num, ] <- T.n[[t]][z.num, ]  # bottom
        T.n[[t + 1]][, 1] <- T.n[[t]][, 1]          # left
        T.n[[t + 1]][, x.num] <- T.n[[t]][, x.num]  # right
        # T.n[[t + 1]][, 1] <- 1500 + (0.5 * -z[[i]] / 1000) + 273.15      # left
        # T.n[[t + 1]][, x.num] <- 1500 + (0.5 * -z[[i]] / 1000) + 273.15  # right
        
        # Calculate T in the interior of the model
        T.n[[t + 1]][i, j] <- T.n[[t]][i, j] +
          
          # conduction term
          ((Kappa[i, j] * dt) * (((T.n[[t]][(i + 1), j] - (2 * T.n[[t]][i, j]) + T.n[[t]][(i - 1), j]) / (dx ^ 2)) + ((T.n[[t]][i, (j + 1)] - (2 * T.n[[t]][i, j]) + T.n[[t]][i, (j - 1)]) / (dz ^ 2) ))) +
          
          # heat production term
          ((A[i, j] * dt) / (rho[i, j] * Cp[i, j])) +

          # advection/uplift term
          (U[i, j] * dt * ((T.n[[t]][i, (j + 1)] - T.n[[t]][i, j]) / dz))
      }
    }
  }
```

### Results after `r eval(n + m)` iterations

Maximum temperature change between model run n = `r (n + m)` (time = `r round(((n + m) * dt * 3.17098e-8 / 1000000), digits = 2)` Myr) and initial condition (time = 0 Myr):  
`r  round(((max(T.n[[length(T.n)]] - T.n[[1]])) - 273.15), digits = 2)` °C   

```{r PlotModel_heatmap_2}
# Plot model results after n iterations.
  plot_ly(x = (x / 1000),
          y = (z / 1000), 
          z = T.n[[length(T.n)]] - 273.15,
          type = "heatmap",
          colors = "Greys") %>%
          # colors = rev(brewer.pal(9,"RdYlBu"))) %>%
    colorbar(title = "Temperature (°C)",
             limits = c(1450, 1750))  %>%
  layout(xaxis = list(title = "Distance (km)"),
         yaxis = list(title = "Depth (km)",
                      autorange = "reversed"))
```

```{r Model_SetIterations_3}
# Set a number of time steps.
  o <- 100  # 100 iterations is about 5 myr

# Solve the finite difference equation for n time steps.
  
  for (t in n:(n + m + o)) {
    
    # Initialize T.n[[t + 1]]
    T.n[[t + 1]] <- matrix(data = NA,
                           nrow = z.num,
                           ncol = x.num)
    
    # Calculate T for interior of T.n[[t + 1]] matrix.
    for (i in 2:(x.num - 1)) {
      for (j in 2:(z.num - 1)) {
            
        # Set constant T border.
        T.n[[t + 1]][1, ] <- T.n[[t]][1, ]          # top
        T.n[[t + 1]][z.num, ] <- T.n[[t]][z.num, ]  # bottom
        T.n[[t + 1]][, 1] <- T.n[[t]][, 1]          # left
        T.n[[t + 1]][, x.num] <- T.n[[t]][, x.num]  # right
        # T.n[[t + 1]][, 1] <- 1500 + (0.5 * -z[[i]] / 1000) + 273.15      # left
        # T.n[[t + 1]][, x.num] <- 1500 + (0.5 * -z[[i]] / 1000) + 273.15  # right
        
        # Calculate T in the interior of the model
        T.n[[t + 1]][i, j] <- T.n[[t]][i, j] +
          
          # conduction term
          ((Kappa[i, j] * dt) * (((T.n[[t]][(i + 1), j] - (2 * T.n[[t]][i, j]) + T.n[[t]][(i - 1), j]) / (dx ^ 2)) + ((T.n[[t]][i, (j + 1)] - (2 * T.n[[t]][i, j]) + T.n[[t]][i, (j - 1)]) / (dz ^ 2) ))) +
          
          # heat production term
          ((A[i, j] * dt) / (rho[i, j] * Cp[i, j])) +

          # advection/uplift term
          (U[i, j] * dt * ((T.n[[t]][i, (j + 1)] - T.n[[t]][i, j]) / dz))
      }
    }
  }
```

### Results after `r eval(n + m + o)` iterations

Maximum temperature change between model run n = `r (n + m + o)` (time = `r round(((n + m + o) * dt * 3.17098e-8 / 1000000), digits = 2)` Myr) and initial condition (time = 0 Myr):  
`r  round(((max(T.n[[length(T.n)]] - T.n[[1]])) - 273.15), digits = 2)` °C   

```{r PlotModel_heatmap_3}
# Plot model results after n iterations.
  plot_ly(x = (x / 1000),
          y = (z / 1000), 
          z = T.n[[length(T.n)]] - 273.15,
          type = "heatmap",
          colors = "Greys") %>%
          # colors = rev(brewer.pal(9,"RdYlBu"))) %>%
    colorbar(title = "Temperature (°C)",
             limits = c(1450, 1750))  %>%
  layout(xaxis = list(title = "Distance (km)"),
         yaxis = list(title = "Depth (km)",
                      autorange = "reversed"))
```
